---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tidybrreg

<!-- badges: start -->
[![R-CMD-check](https://github.com/sondreskarsten/tidybrreg/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sondreskarsten/tidybrreg/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

Tidy R interface to Norway's [Central Coordinating Register for Legal Entities](https://www.brreg.no/en/) (Enhetsregisteret), maintained by the Brønnøysund Register Centre. The register contains approximately 1 million active entities. Data is freely available under the [Norwegian Licence for Open Government Data (NLOD 2.0)](https://data.norge.no/nlod/en/2.0).

## Installation

```{r install, eval = FALSE}
# install.packages("pak")
pak::pak("sondreskarsten/tidybrreg")
```

## Entity lookup

```{r entity, eval = FALSE}
library(tidybrreg)

brreg_entity("923609016")
#> # A tibble: 1 × 65
#>   org_nr    name        legal_form founding_date employees nace_1
#>   <chr>     <chr>       <chr>      <date>            <int> <chr>
#> 1 923609016 EQUINOR ASA ASA        1972-09-18        21408 06.100
```

Codes by default. Translate to English with `type = "label"` or `brreg_label()`:

```{r label, eval = FALSE}
brreg_entity("923609016", type = "label")
#> legal_form: "Public limited company"
#> nace_1:     "Extraction of crude petroleum"

# Keep both codes and labels (eurostat pattern)
brreg_entity("923609016") |>
  brreg_label(code = c("legal_form", "nace_1"))
#> legal_form_code: "ASA"
#> legal_form:      "Public limited company"
#> nace_1_code:     "06.100"
#> nace_1:          "Extraction of crude petroleum"
```

## Search

```{r search, eval = FALSE}
brreg_search(legal_form = "AS", municipality_code = "0301",
             min_employees = 500, max_results = 10)
```

## Board members and officers

```{r roles, eval = FALSE}
brreg_roles("923609016")
#> # A tibble: 16 × 14
#>   org_nr    role_group         role                first_name last_name
#>   <chr>     <chr>              <chr>               <chr>      <chr>
#> 1 923609016 Management         CEO / Managing Dir… Anders     Opedal
#> 2 923609016 Board of Directors Chair of the Board  Jon Erik   Reinhardsen
#> …

brreg_roles("923609016") |> brreg_board_summary()
#> board_size: 14 | n_chair: 1 | has_ceo: TRUE | has_auditor: TRUE
```

## Bulk download

```{r download, eval = FALSE}
# Full register (~1M rows, 145 MB gzipped)
entities <- brreg_download()

# Lazy Arrow Table (not loaded into memory)
entities_arrow <- brreg_download(type_output = "arrow")

# Incremental updates (CDC)
brreg_updates(since = Sys.Date() - 7)
```

## Validate organization numbers

```{r validate}
library(tidybrreg)
brreg_validate(c("923609016", "984851006", "123456789"))
```

## Design

**Codes by default, labels on demand.** Functions return Norwegian codes. `brreg_label()` translates to English via bundled reference data or live [SSB Klass API](https://data.ssb.no/api/klass/v1/) lookups (`brreg_label(x, refresh = TRUE)`).

**Data-driven column dictionary.** `field_dict` maps Norwegian API names to English. Unknown API fields pass through with auto-generated snake_case names—new fields added by brreg are never silently dropped.

**Separate functions for separate backends.** `brreg_search()` uses the JSON API (filtered, max 10,000). `brreg_download()` uses the bulk CSV (always full register). Both return the same column schema via `field_dict`.

## Functions

| Function | Description |
|---|---|
| `brreg_entity()` | Single entity by organization number |
| `brreg_search()` | Filtered search (name, legal form, industry, geography) |
| `brreg_roles()` | Board members, officers, auditors |
| `brreg_board_summary()` | Board-level covariates from role data |
| `brreg_download()` | Full register bulk download (~1M rows) |
| `brreg_updates()` | Incremental change stream (CDC) |
| `brreg_label()` | Translate codes to English descriptions |
| `brreg_validate()` | Organization number validation (modulus-11) |
| `get_brreg_dic()` | Fetch/cache NACE or sector dictionaries |

## Reference datasets

| Dataset | Description |
|---|---|
| `field_dict` | Column name mapping (49 Norwegian → English) |
| `legal_forms` | 44 legal form codes with English translations |
| `role_types` | 18 role codes with English translations |
| `role_groups` | 15 role group codes with English translations |

## License

MIT. Data from Enhetsregisteret is available under [NLOD 2.0](https://data.norge.no/nlod/en/2.0).
