---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tidybrreg

<!-- badges: start -->
[![R-CMD-check](https://github.com/sondreskarsten/tidybrreg/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sondreskarsten/tidybrreg/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

Tidy R interface to Norway's [Central Coordinating Register for Legal Entities](https://www.brreg.no/en/) (Enhetsregisteret), maintained by the Brønnøysund Register Centre. The register contains approximately 1 million active entities. Data is freely available under the [Norwegian Licence for Open Government Data (NLOD 2.0)](https://data.norge.no/nlod/en/2.0).

## Installation

```{r install, eval = FALSE}
# install.packages("pak")
pak::pak("sondreskarsten/tidybrreg")

# For snapshot/panel features, also install a parquet backend:
install.packages("nanoparquet")  # lightweight
# OR
install.packages("arrow")       # full-featured (lazy queries)
```

## Entity lookup

```{r entity, eval = FALSE}
library(tidybrreg)

brreg_entity("923609016")
#> # A tibble: 1 x 65
#>   org_nr    name        legal_form founding_date employees nace_1
#>   <chr>     <chr>       <chr>      <date>            <int> <chr>
#> 1 923609016 EQUINOR ASA ASA        1972-09-18        21408 06.100
```

Codes by default. Translate with `type = "label"` or `brreg_label()`:

```{r label, eval = FALSE}
# Inline labelling
brreg_entity("923609016", type = "label")

# Keep both codes and labels (eurostat pattern)
brreg_entity("923609016") |>
  brreg_label(code = c("legal_form", "nace_1"))
#> legal_form_code: "ASA"
#> legal_form:      "Public limited company"
#> nace_1_code:     "06.100"
#> nace_1:          "Extraction of crude petroleum"
```

## Search

```{r search, eval = FALSE}
brreg_search(legal_form = "AS", municipality_code = "0301",
             min_employees = 500, max_results = 10)
```

## Board members and officers

```{r roles, eval = FALSE}
brreg_roles("923609016")
brreg_roles("923609016") |> brreg_board_summary()
```

## Snapshot engine

Save dated bulk downloads as Hive-partitioned Parquet for panel construction:

```{r snapshot, eval = FALSE}
# Save today's full register (~145 MB download, ~90 MB parquet)
brreg_snapshot()

# Import historical CSVs
brreg_import("enheter_2024-12-31.csv.gz", snapshot_date = "2024-12-31")

# List available snapshots
brreg_snapshots()
#> # A tibble: 3 x 3
#>   snapshot_date file_size path
#>   <date>            <dbl> <chr>
#> 1 2024-01-01     87654321 ~/.local/share/R/tidybrreg/enheter/...
#> 2 2024-07-01     88123456 ...
#> 3 2025-01-01     89012345 ...
```

## Panel construction

Build firm x period panels from accumulated snapshots:

```{r panel, eval = FALSE}
# Annual panel with selected variables
panel <- brreg_panel(
  frequency = "year",
  cols = c("employees", "nace_1", "legal_form", "municipality_code")
)

# Monthly panel
panel_m <- brreg_panel("month", cols = c("employees"))

# Detect entries, exits, and field changes between two dates
events <- brreg_events("2024-01-01", "2025-01-01")
events |> dplyr::count(event_type)
#>   event_type     n
#> 1 change     52314
#> 2 entry       8721
#> 3 exit        3102

# Aggregate time series
brreg_series("count", by = "legal_form", frequency = "year")
```

## Harmonization

Handle municipality boundary reforms and NACE reclassifications:

```{r harmonize, eval = FALSE}
panel |> brreg_harmonize_kommune(target_date = "2024-01-01")
panel |> brreg_harmonize_nace(from = "SN2007", to = "SN2025")
```

## Governance research

```{r governance, eval = FALSE}
# Director interlock network (requires tidygraph)
net <- brreg_board_network(c("923609016", "984851006"))

# Firm survival data (for survival::Surv())
surv <- brreg_search(legal_form = "AS", max_results = 1000) |>
  brreg_survival_data()
```

## Validate organization numbers

```{r validate}
library(tidybrreg)
brreg_validate(c("923609016", "984851006", "123456789"))
```

## Design

**Codes by default, labels on demand.** Functions return Norwegian codes. `brreg_label()` translates to English via bundled reference data or live [SSB Klass API](https://data.ssb.no/api/klass/v1/) lookups.

**Data-driven column dictionary.** `field_dict` maps Norwegian API names to English. Unknown API fields pass through with auto-generated snake_case names.

**Separate backends for separate scales.** `brreg_search()` uses the JSON API (filtered, max 10,000). `brreg_download()` / `brreg_snapshot()` use the bulk CSV (always full register). Both produce the same column schema via `field_dict`.

**Snapshot-based temporal database.** `brreg_snapshot()` accumulates Hive-partitioned Parquet files. `brreg_panel()` queries them lazily via Arrow with partition pruning and projection pushdown.

## Functions

| Function | Description |
|---|---|
| `brreg_entity()` | Single entity by organization number |
| `brreg_search()` | Filtered search (name, legal form, industry, geography) |
| `brreg_roles()` | Board members, officers, auditors |
| `brreg_board_summary()` | Board-level covariates from role data |
| `brreg_download()` | Full register bulk download (~1M rows) |
| `brreg_updates()` | Incremental change stream (CDC) |
| `brreg_label()` | Translate codes to English descriptions |
| `brreg_validate()` | Organization number validation (modulus-11) |
| `get_brreg_dic()` | Fetch/cache NACE or sector dictionaries |
| `brreg_snapshot()` | Save dated bulk download as Parquet partition |
| `brreg_import()` | Import historical CSV as snapshot partition |
| `brreg_snapshots()` | List available snapshots |
| `brreg_open()` | Open snapshot store as lazy Arrow Dataset |
| `brreg_cleanup()` | Remove old snapshot partitions |
| `brreg_panel()` | Firm x period panel from snapshots |
| `brreg_events()` | Snapshot diff: entries, exits, field changes |
| `brreg_series()` | Aggregate time series from snapshots |
| `brreg_harmonize_kommune()` | Municipality code harmonization |
| `brreg_harmonize_nace()` | NACE code harmonization (SN2007/SN2025) |
| `brreg_board_network()` | Director interlock network (tidygraph) |
| `brreg_survival_data()` | Firm survival data preparation |
| `brreg_data_dir()` | Snapshot store location |

## License

MIT. Data from Enhetsregisteret is available under [NLOD 2.0](https://data.norge.no/nlod/en/2.0).
